---
title: "title"
author: "Quentin D. Read"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

What is this course? It is a crash course in **brms**: a brief and practical introduction to fitting Bayesian multilevel models in R and Stan using the [**brms**](https://paul-buerkner.github.io/brms/) package (**B**ayesian **R**egression **M**odels using **S**tan). Because this is intended as a practical skills course, I will only give a quick and surface-level description of Bayesian analysis and how it can be used for multilevel regression models.

## What is Bayesian analysis?

Laplace and Reverend Bayes independently invneted it

## What is Stan?

Andrew Gelman and colleagues developed

## What is brms?

Paul Buerkner developed

## Why use it?

- Philosophical reasons
- Math reasons

# Installing brms

```{r load packages}

```

```{r ggplot theme, include = FALSE}
windowsFonts(`fgbook` = windowsFont("Franklin Gothic Book"))
theme_set(
  theme_bw(base_family = 'fgbook') + 
    theme(panel.grid = element_blank(),
          strip.background = element_blank())
)
```

# Resources

Here are some links to tutorials and lessons for learning basic Bayesian statistics with **brms**.

- [Building a multilevel model in brms](https://www.rensvandeschoot.com/tutorials/brms-started/) by Rens van de Schoot
- [An introduction to Bayesian multilevel models using brms](https://www.barelysignificant.com/phd_thesis/appendix-brms.html) by Ladislas Nalborczyk

Paul Buerkner published two papers on **brms**:

- [Buerkner 2017, Journal of Statistical Software](https://www.jstatsoft.org/article/view/v080i01)
- [Buerkner 2018, The R Journal](https://journal.r-project.org/archive/2018/RJ-2018-017/index.html)

# Alternatives

- rstanarm
- MCMCglmm
- PyStan (other platforms for Stan)
- CmdStan